Parsing Postfix log files

### 2009-03-26 doc/logparser-complications.tex
Corrections from proofreading, with lots more to come.
More corrections from proofreading.

\section{Complications Encountered}

@@

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the \daemon{smtpd} \acronym{pid}, then transition to
being correlated by the queueid.  This is relatively minor, but does
require:

@@

    \item Two state tables to hold the data structure for each connection
        and mail.

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by \acronym{pid} or queueid.

@@

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log lines will not be correlated by
\acronym{pid} any more, its queueid will be used instead (except when
timeouts occur during the data phase \sectionref{timeouts during data
phase}).  If the original connection has any useful data (e.g.\ rejections)
it will be saved to the database when the client disconnects.  One unsolved
difficulty is distinguishing between different groups of rejections, e.g.\
when dealing with the following sequence:

@@

Mails sent to local addresses are not always delivered directly to a
mailbox: sometimes they are sent to and accepted for a local address, but
need to be delivered to one or more remote addresses due to aliasing.  When
this occurs, a child mail will be injected into the Postfix queue, but
without the explicit logging that mails injected by \daemon{smtpd} or
\daemon{postdrop} have.  Thus the source of the mail is not immediately
discernible from the log line in which the mail's queueid first appears:
from a strictly chronological reading of the log lines it usually appears
as if the child mail has been created without a source.  Subsequently the
parent mail will log the creation of the child mail, e.g.\ parent mail
3FF7C4317 creates child mail 56F5B43FD\@:

@@

Unfortunately, while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition, the first log line belonging to the child mail (the example log
line above belongs to the parent mail) is logged by \daemon{qmgr} or
\daemon{cleanup}, so the order also depends on how soon they process the
new mail.

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs the creation log line!  Thus, mails
flagged as coming from an unknown origin cannot be entered into the
database when their final log line is processed; instead they must be
marked as ready for entry and subsequently entered once their origin has
been identified.  The crux of this complication is that re-injected mails
appear in the log files without explicit logging indicating their source.

@@

    \item If a mail is tagged with the unknown origin flag it is assumed to
        be a child mail whose parent has not yet been identified.  The mail
        is tagged as ready to be entered in the database, but entry is
        postponed until the parent is identified.  The child mail should
        not have any subsequent log lines: only its parent should refer to
        it.

    \item If the mail is a child mail with an identified parent it is
        entered in the database as usual, then removed from its parent's
        list of children.  If this child is the last mail on that list, and
        the parent has already been entered in the database, the parent
        will be removed from the state tables.

    \item If the mail is a parent it is entered in the database as usual,
        because there will be no further log lines for it.  There may be
        child mails waiting to be entered in the database; these are
        entered as normal, and removed from the state tables.  If there are
        more child mails, the parent's removal from the state tables will
        be postponed until the last child has been entered.

@@

Postfix 2.2.x (and presumably previous versions) does not generate a log
line when it generates a bounce notification; suddenly there will be log
entries for a mail that lacks an obvious source.  There are similarities to
the problem of identifying re-injected mails discussed in
\sectionref{tracking re-injected mail}, but unlike the solution described
therein bounce notifications do not eventually have a log line that
identifies their source.  Heuristics must be used to identify bounce
notifications, and those heuristics are:

@@

    \item The message-id has a specific format: \newline{}
        \tab{} \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing.  In the latter case the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format (e.g.\ if it was generated by
        Postfix on this or another server) it will not match the queueid of
        the mail.

@@

Some mail clients behave unexpectedly during the \acronym{SMTP} dialogue:
the client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.  Microsoft Outlook is one
client that behaves in this fashion; other clients may act in a similar
way.  An example dialogue exhibiting this behaviour is presented below
(lines starting with a three digit number are sent by the server, the other
lines are sent by the client):

@@

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify when this
behaviour occurs.  In this case a list of all mails accepted during a
connection is saved in the connection state, and the accepted mails are
examined when the disconnection action is executed.  Each mail is checked
for the following:

@@

    \item Was the second result processed by the \texttt{CLONE} action?
        The first two \daemon{smtpd} log lines will be a connection log
        line and a mail acceptance log line; Postfix logs acceptance as
        soon as the first recipient has been accepted.

@@

    \item Does the queueid exist in the state tables?  If not it cannot be
        an aborted delivery attempt.

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \acronym{IP} address, but the database
schema requires many more fields be populated (see \sectionref{connections
table} and \sectionref{results table}).  This heuristic is quite
restrictive, and it appears there is little scope for false positives; if
there are any false positives there will be warnings when the next log line
for that mail is parsed.  False negatives are less likely to be detected:
there may be queueid clashes (and warnings) if mails remain in the state
tables after they should have been removed, otherwise the only way to
detect false negatives is to examine the state tables after each parsing
run.

This check is performed in the DISCONNECT action; it requires support in
the CLONE action where a list of cloned connections is maintained.

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser (\sectionref{timeouts during data phase}).  Sadly, Postfix does not
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

    \item Is the mail missing its \daemon{cleanup} log line?  Every mail
        which passes through Postfix will have a \daemon{cleanup} line;
        lack of a \daemon{cleanup} line is a sure sign the mail did not
        make it too far.

    \item Were there three or more \daemon{smtpd} log lines for the mail?
        There should be a connection log line and a mail accepted log line,
        followed by one or more rejection log lines.

    \item Is the last \daemon{smtpd} log line a rejection line?

@@

If all checks are successful then the mail is assumed to have been
discarded when the client disconnected.  There will be no further entries
logged for such mails, so without identifying and entering them in the
database immediately they accumulate in the state table and will cause
clashes if the queueid is reused.

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes there is a timeout or the
connection is lost during the DATA phase;\footnote{For brevity's sake
\textit{timeout\/} will be used throughout this section, but everything
applies equally to lost connections.} when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded and saved with the connection instead.  To deal properly with
timeouts the TIMEOUT action does the following:

@@

    \item Record the timeout and associated data in the connection's
        results.

@@

    \item If one or more recipients have been accepted, Postfix will have
        allocated a queueid for the incoming mail, and there may be a mail
        in the state tables that needs to be discarded.  The timeout may
        have interrupted transfer of an accepted delivery attempt, or it
        may have occurred after a mail delivery attempt was rejected.  To
        distinguish between the two cases the action compares the timestamp
        of the last accepted mail against the timestamp of the last log
        line from \daemon{smtpd} for that connection; this depends on the
        CLONE action keeping a list of all mails accepted for each
        connection.  If the mail acceptance timestamp is later than the
        timeout timestamp, the timeout applies to the just-accepted mail,
        which will be removed from the state tables, without entering it in
        the database.  If the \daemon{smtpd} timestamp is later there was a
        rejection between the accepted mail and the timeout: the action
        assumes that the timeout applies to the rejected delivery attempt
        and finishes.  This assumption is not necessarily correct, because
        Postfix may have accepted an earlier recipient and rejected a later
        one, in which case the timeout applies to the accepted mail, which
        should be discarded; this has not been a problem in practice.
        Processing timeouts is further complicated by the presence of out
        of order \daemon{cleanup} log lines: see \sectionref{discarding
        cleanup log lines} for details.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} line for a mail being accepted is logged after the timeout
line; parsing this line causes the MAIL\_QUEUED action to create a new
state table entry for the queueid in the log line.  This is incorrect
because the line actually belongs to the mail that has just been discarded;
the next log line for that queueid will be seen when the queueid is reused
for a different mail, causing a queueid clash and the appropriate warning.

When the \daemon{cleanup} line is still pending during the TIMEOUT action,
the action updates a global list of queueids, adding the queueid and the
timestamp from the log line.  When the next \daemon{cleanup} line is parsed
for that queueid the list will be checked (during the MAIL\_QUEUED action),
and the log line will be deemed part of the mail where the timeout occurred
and discarded if it meets the following requirements:

@@

    \item The queueid must not have been reused yet, i.e.\ it does not have
        an entry in the state tables.

    \item The timestamp of the \daemon{cleanup} log line must be within ten
        minutes of the mail acceptance timestamp.  Timeouts happen after
        five minutes, but some data may have been transferred slowly
        (perhaps because either the client or server is suffering from
        network congestion or rate limiting), and empirical evidence shows
        that ten minutes is not unreasonable; hopefully it is a good
        compromise between false positives (log lines incorrectly
        discarded) and false negatives (new state table entries incorrectly
        created).

@@

The next \daemon{cleanup} line must meet the criteria above for it to be
discarded because some, but not all connections where a timeout occurs will
have an associated \daemon{cleanup} line logged; if the algorithm blindly
discarded the next \daemon{cleanup} line after a timeout it would sometimes
be mistaken.  When the next \daemon{pickup} line containing that queueid is
parsed the queueid will be removed from the cache of timeout queueids,
regardless of whether it meets the criteria above.

This complication is handled by the TIMEOUT and MAIL\_QUEUED actions.

@@

>>>>

\section{Complications Encountered While Writing \parsernamelong{}}

@@

\label{queueid vs pid}

A delivery attempt lacks a queueid until one recipient has been accepted,
so log lines must first be correlated by \daemon{smtpd} \acronym{pid}, then
transition to being correlated by their queueid.  This is relatively minor,
but does require:

@@

    \item Two state tables to hold data structures.

    \item Most importantly: every section of code must know which state
        table it should query.

@@

Multiple independent mails may be delivered across one connection: this
requires \parsername{} to clone the connection's data as soon as a mail is
accepted, so that subsequent mails will not overwrite each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once a mail
has been accepted its log lines will not be correlated by \acronym{pid} any
more, its queueid will be used instead, as described in \sectionref{queueid
vs pid}.  If the original connection has any useful data (e.g.\ rejections)
it will be saved to the database when the client disconnects.  One unsolved
difficulty is distinguishing between different groups of rejections, e.g.\
when dealing with the following sequence:

@@

Mails sent to local addresses are not always delivered directly to a
mailbox: sometimes they are sent to and accepted for a local address, but
need to be delivered to one or more remote addresses due to aliasing.  When
this occurs, a child mail will be injected into the Postfix queue, but
without the explicit logging that mails injected by \daemon{smtpd} or
\daemon{postdrop} have.  Thus the source of the mail is not immediately
discernible from the log line in which the mail's queueid first appears:
from a strictly chronological reading of the log lines it usually appears
as if the child mail has been created without a source.  Subsequently the
parent mail will log the creation of the child mail, e.g.\ parent mail
\texttt{3FF7C4317} creates child mail \texttt{56F5B43FD}\@:

@@

Unfortunately, while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition, the first log line belonging to the child mail (the example log
line above belongs to the parent mail) is logged by either \daemon{qmgr} or
\daemon{cleanup}, so the order also depends on how soon they process the
new mail.

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, nor can \parsername{} warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  The crux of this complication is
that re-injected mails appear in the log files without explicit logging
indicating their source.

@@

    \item If a mail is tagged with the unknown origin flag, it is assumed
        to be a child mail whose parent has not yet been identified.  The
        mail is tagged as ready to be entered in the database, but entry is
        postponed until the parent is identified.  The child mail should
        not have any subsequent log lines: only its parent will refer to
        it.

    \item If the mail is a child mail whose parent has been identified, it
        is entered in the database as usual, then removed from its parent's
        list of children.  If this child is the last mail on that list, and
        the parent has already been entered in the database, the parent
        will be removed from the state tables.

    \item If the mail is a parent, it is entered in the database as usual
        because there will be no further log lines for it.  There may be
        child mails waiting to be entered in the database; these are
        entered as normal, and removed from the state tables.  If there are
        incomplete child mails, the parent's removal from the state tables
        will be postponed until the last child has been entered.

@@

Postfix 2.2.x (and presumably previous versions) does not generate a log
line when it generates a bounce notification; suddenly there will be log
entries for a mail that lacks an obvious source.  There are similarities to
the problem of identifying re-injected mails discussed in
\sectionref{tracking re-injected mail}, but unlike the solution described
therein bounce notifications do not eventually have a log line that
identifies their source.  Heuristics must be used to identify bounce
notifications:

@@

    \item The message-id has a specific format: \newline{}
        \tab{} \texttt{YYYYMMDDhhmmss.queueid@server\_hostname} \newline{}
        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes a new bounce notification from
        a bounce notification that is being re-injected as a result of
        aliasing.  In the latter case the message-id will be unchanged from
        the original bounce notification, and so even if it happens to be
        in the correct format, e.g.\ if it was generated by Postfix on this
        or another server, it will not match the queueid of the mail.

@@

Some mail clients behave strangely during the \acronym{SMTP} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient that
it continues with until delivery is complete.  Microsoft Outlook is one
client that behaves in this fashion; other clients may act in a similar
way.  An example dialogue exhibiting this behaviour is presented below
(lines starting with a three digit number are sent by the server, the other
lines are sent by the client):

@@

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify when this
behaviour occurs.  A list of all mails accepted during a connection is
saved in the connection's state table entry, and the accepted mails are
examined when the disconnection action is invoked.  Each mail is checked
for the following:

@@

    \item Was the second result processed by the \texttt{CLONE} action?
        The first two \daemon{smtpd} log lines will be a connection log
        line and a mail acceptance log line.

@@

    \item Can the mail be found in the state tables?  If not it cannot be
        an aborted delivery attempt.

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \acronym{IP} address, but the database
schema requires many more fields be populated (see \sectionref{connections
table} and \sectionref{results table}).  These heuristics are quite
restrictive, and it appears there is little scope for false positives; if
there are any false positives there will also be warnings when the next log
line for that mail is parsed.  False negatives are less likely to be
detected: there may be queueid clashes (and warnings) if mails remain in
the state tables after they should have been removed, otherwise the only
way to detect false negatives is to examine the state tables after each
parsing run.

This check is performed in the DISCONNECT action; it requires support in
the CLONE action where the list of accepted mails is maintained.

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser, e.g.\ \sectionref{timeouts during data phase}.  Sadly, Postfix does
not log a message saying the mail has been discarded, as should be expected
by now.  The checks to identify this happening are:

@@

    \item Is the mail missing its \daemon{cleanup} log line?  Every mail
        which passes normally through Postfix will have a \daemon{cleanup}
        line; lack of one is a sure sign the mail did not make it too far.

    \item Were there three or more \daemon{smtpd} log lines for the mail?
        There should be a connection log line and a mail acceptance log
        line, followed by one or more delivery rejected log lines.

@@

If both checks are successful then the mail is assumed to have been
discarded when the client disconnected.  There will be no further log lines
for such mails, so without identifying and entering them in the database
immediately they accumulate in the state table and will cause clashes if
the queueid is reused.

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes there is a timeout or the
connection is lost during the DATA phase;\footnote{For the sake of brevity
\textit{timeout\/} will be used throughout this section, but everything
applies equally to lost connections.} when this occurs Postfix will discard
the mail and \parsername{} needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded and saved with the connection instead.  To deal properly with
timeouts the TIMEOUT action does the following:

@@

    \item Record the timeout and captured data in the connection's results.

@@

    \item If one or more recipients have been accepted, Postfix will have
        allocated a queueid for the incoming mail, and there may be a mail
        in the state tables that needs to be discarded.  The timeout may
        have interrupted transfer of an accepted delivery attempt, or it
        may have occurred after a mail delivery attempt was rejected.

        XXX THE REMAINDER OF THIS DOES NOT MAKE SENSE\@; CHECK AND CORRECT
        IT\@.  To distinguish between the two cases the action compares the
        timestamp of the last accepted mail against the timestamp of the
        last log line from \daemon{smtpd} for that connection; this depends
        on the CLONE action keeping a list of all mails accepted for each
        connection.  If the mail acceptance timestamp is later than the
        timeout timestamp, the timeout applies to the just-accepted mail,
        which will be removed from the state tables, without entering it in
        the database.  If the \daemon{smtpd} timestamp is later there was a
        rejection between the accepted mail and the timeout: the action
        assumes that the timeout applies to the rejected delivery attempt
        and finishes.  This assumption is not necessarily correct, because
        Postfix may have accepted an earlier recipient and rejected a later
        one, in which case the timeout applies to the accepted mail, which
        should be discarded; this has not been a problem in practice.
        Processing timeouts is further complicated by the presence of out
        of order \daemon{cleanup} log lines: see \sectionref{discarding
        cleanup log lines} for details.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} log line for a mail being accepted is logged after the
timeout log line, by which time \parsername{} has discarded the mail;
parsing the \daemon{cleanup} log line causes the CLEANUP\_PROCESSING action
to create a new state table entry, to help deal with re-injected mails
(\sectionref{Re-injected mails}).  This is incorrect because the log line
actually belongs to the mail that has just been discarded; if the queueid
is reused there will be a queueid clash, otherwise the new mail will just
remain in the state tables.

During the TIMEOUT action, if the \daemon{cleanup} line is still pending,
the action updates the \texttt{timeout\_queueids} state table, adding the
queueid and the timestamp from the log line.  When the next
\daemon{cleanup} line is parsed for that queueid, the list will be checked
and if the log line meets the following requirements it will be deemed part
of the mail where the timeout occurred and discarded.

@@

    \item The queueid must not have been reused yet, i.e.\ it does not have
        a mail in the state tables.

    \item The timestamp of the \daemon{cleanup} log line must be within ten
        minutes of the mail acceptance timestamp.  Timeouts happen after
        five minutes, but some data may have been transferred slowly, and
        empirical evidence shows that ten minutes is not unreasonable;
        hopefully it is a good compromise between false positives (log
        lines incorrectly discarded) and false negatives (new state table
        entries incorrectly created).

@@

The \daemon{cleanup} log line must meet the criteria above for it to be
discarded because some, but not all, connections where a timeout occurs
will have an associated \daemon{cleanup} log line; if the
CLEANUP\_PROCESSING action blindly discarded the next \daemon{cleanup} log
line after a timeout it would sometimes be mistaken.  When the next
\daemon{pickup} log line containing that queueid is processed the queueid
will be removed from the cache of timeout queueids, regardless of whether
it meets the criteria above.

This complication is handled by the CLEANUP\_PROCESSING and TIMEOUT
actions.

@@

\newpage{} %XXX VARIOREF PROBLEM
